---
title: "Intro to Clinical Research"
author: "Madelyn Houser"
date: "9/29/2020"
output: 
  word_document:
    toc: yes
  html_document:
    toc: yes
---

```{r version}
# This will document the version of R that I am using for this analysis.
version
```

```{r packages}
# This will load the packages needed for this analysis and document the version of each that I'm using

# If you have not previously installed any of these packages, install them using the function
## install.packages("PACKAGENAME")

library("Hmisc")
packageVersion("Hmisc")
library("tidyverse")
packageVersion("tidyverse")
library("desctable")
packageVersion("desctable")
library("ggpubr")
packageVersion("ggpubr")
```
  
# Import and merge data
  
```{r import}
demo <- sasxport.get("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/NHANES/DEMO_J.XPT")
cbc <- sasxport.get("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/NHANES/CBC_J.XPT")
hepc <- sasxport.get("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/NHANES/HEPC_J.XPT")
```
```{r select}
# Select only the variables of interest from the data sets
demo <- demo %>% select(seqn, riagendr, ridageyr, ridagemn)
```
  
```{r merge}
# Merge into one data set, keeping only rows with matching identifiers in demographic data set
df <- merge(x=demo, y=cbc, by="seqn", all.x=TRUE)
df <- merge(x=df, y=hepc, by="seqn", all.x=TRUE)
```

  
  
# Explore and Summarize data
  
```{r explore}
# Report class of the data set
class(df)

# This will provide some basic information on each variable in the data set
describe(df)
```
  
# Data cleanup and wrangling
  
Let's check the format of our variables and label the levels of categorical variables and make sure they're classified as factors.
  
```{r label}
# Check class of variable
class(df$riagendr)
# Make variable a factor and define levels and labels of levels
df$riagendr <- factor(df$riagendr, levels=c("1", "2"), labels=c("Male", "Female"))
# Change the name of the variable
df <- rename(df, "Sex" = "riagendr")
# Check the class of the variable again to be sure that the changes were made correctly
class(df$riagendr)

# Check class and change label of another variable
class(df$ridageyr)
label(df$ridageyr) <- "Age (years)"
df <- rename(df, "Age" = "ridageyr")

# Spot check class of lab test variables
class(df$lbxwbcsi)
class(df$lbxlypct)
class(df$lbxnrbc)

# Fix formatting of Hep C RNA variable
class(df$lbxhcr)
df$lbxhcr <- factor(df$lbxhcr, levels=(c("1", "2", "3")), labels=c("Positive", "Negative", "Negative Ab Screening"))
df <- rename(df, "HepC_RNA" = "lbxhcr")
class(df$HepC_RNA)
# Check factor levels of the variable
levels(df$HepC_RNA)

# Rename some other variables
df <- rename(df, "RBCs" = "lbxrbcsi")
df <- rename(df, "Leukocytes" = "lbxwbcsi")
df <- rename(df, "Lymphocytes" = "lbxlypct")
```
  
Let's filter the data so that we're only analyzing data for adults
  
```{r filter}
dfad <- df %>% filter(Age >= 18)

summary(dfad$Age)
```
  
# Visualize data for variables of interest
  
## Table - Categorical Variables by Sex
  
```{r tab1}
# Create a table summarizing some variables of interest by sex
Tab1 <- dfad %>% group_by(Sex) %>% select(Age, HepC_RNA) %>% desctable() %>% datatable()

Tab1
```
  
## Variable graphs
  
```{r plotcatvar}
# Plot counts of each level of a categorical variable
ggplot(data = dfad) + 
  geom_bar(mapping = (aes(x = Sex, fill=Sex)))

ggplot(data = dfad) + 
  geom_bar(mapping = (aes(x = HepC_RNA, fill=HepC_RNA)))

# Plot Age split by Sex
ggplot(data = dfad) +
  geom_boxplot(mapping=aes(x=Sex, y=Age, fill=Sex))
```
  
```{r plotconvar}
# Create a boxplot of RBCs data split by Sex
ggplot(data=dfad, mapping=aes(x=Sex, y=RBCs, fill=Sex)) +
  geom_boxplot()

# Create a boxplot of RBCs data split by HepC_RNA
ggplot(data=dfad, mapping=aes(x=HepC_RNA, y=RBCs, color=HepC_RNA)) +
  geom_boxplot()

# Repeat with other continuous variables
ggplot(data=dfad, mapping=aes(x=Sex, y=Leukocytes, fill=Sex)) +
  geom_boxplot()

ggplot(data=dfad, mapping=aes(x=HepC_RNA, y=Leukocytes, color=HepC_RNA)) +
  geom_boxplot()

ggplot(data=dfad, mapping=aes(x=Sex, y=Lymphocytes, fill=Sex)) +
  geom_boxplot()

ggplot(data=dfad, mapping=aes(x=HepC_RNA, y=Lymphocytes, color=HepC_RNA)) +
  geom_boxplot()

# Let's make sure we have enough data to work with within our HepC positive group if we split it by sex

# Create a data set with only subjects positive for HepC RNA
dfadpos <- dfad %>% filter(HepC_RNA == "Positive")

ggplot(data=dfadpos, mapping=aes(x=Sex, y=RBCs, color=Sex)) +
  geom_jitter()
```
  
#### We observe that there is an extremely high value in the Leukocytes variable data.
  
  
## Check distribution of continuous variables
  
### Density and Q-Q plots
  
Let's plot the distribution of some continuous variables.
  
```{r distrib, fig.height=2, fig.width=4}
# Create an object consisting of a plot of the distribution of the RBCs variable
denplot <- ggdensity(df$RBCs, xlab = "Red Blood Cells (million cells/uL)")
# Create an object consisting of a Q-Q plot of the RBCs variable
qqplot <- ggqqplot(df$RBCs, xlab = "Red Blood Cells (million cells/uL)")
# Show the plots side by side
ggarrange(denplot, qqplot, ncol=2, nrow=1)


# Repeat for the next variable of interest
denplot <- ggdensity(df$Leukocytes, xlab = "Leukocytes (1000 cells/uL)")
qqplot <- ggqqplot(df$Leukocytes, xlab = "Leukocytes (1000 cells/uL)")

ggarrange(denplot, qqplot, ncol=2, nrow=1)

# Repeat for the next variable of interest
denplot <- ggdensity(df$Lymphocytes, xlab = "Lymphocytes (%)")
qqplot <- ggqqplot(df$Lymphocytes, xlab = "Lymphocytes (%)")

ggarrange(denplot, qqplot, ncol=2, nrow=1)
```
  
#### Results: The distributions of the RBCs, Leukocytes, and Lymphocytes variables appear roughly normal with a few unusually high values.
  
# Remove outlier
  
There is one extremely high value in the Leukocytes variable data that is not at all comparable to the other values. Let's exclude that subject from our analysis.
  
```{r rmoutlier}
# Return the value in column 1 that has the maximum value in the Leukocytes column in the dfad data set
# This relies on indexing, which means referring to a specific position in the dataset
# In R, indexing takes the format: data[row#, column#]
print(dfad[which.max(dfad$Leukocytes),1])

# Remove the row with the outlier (keep all data that do not match the selected ID number in the seqn column)
dfad <- dfad %>% filter(seqn != "102389")
```

  
# Is there a sex difference in Hepatitis C RNA positivity?
  
Let's run a chi-square test to address this question.
  
```{r HepCSex}
# Chi-square test
stats::chisq.test(x=dfad$Sex, y=dfad$HepC_RNA)

# Let's focus on just the positives and negatives in people who got an RNA test
# Let's also just drop the NAs in the HepC_RNA column to simplify graphing
dfadHepC <- dfad %>% drop_na(HepC_RNA) %>% filter(HepC_RNA != "Negative Ab Screening")

# Run the Chi-square test on just the RNA test subjects
stats::chisq.test(x=dfadHepC$Sex, y=dfadHepC$HepC_RNA)
```
  
Let's make a graph to visualize this comparison.
  
```{r HepCSexGraph, fig.height=3, fig.width=3}
ggplot(data=dfadHepC, mapping=aes(x=HepC_RNA, fill=Sex)) +
  geom_bar() +
  xlab("Hepatitis C RNA") +
  ylab("Count")

# Save this plot
ggsave("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/HepCRNA_by_Sex.pdf")
```
  
### Results: HepC RNA positivity by sex
  
#### Positivity in an RNA test for hepatitis C does differ significantly by sex. Males are more likely than females to test positive for hepatitis C RNA.
  
  
  
# Do leukocyte counts differ in subjects who test positive for hepatitis C from those that test negative?
  
Let's run a one-way ANOVA to address this question.
  
```{r leukHepC}
# One-way ANOVA 
res <- aov(Leukocytes ~ HepC_RNA, dfad)
summary(res)
```
  
Let's make a graph to visualize this comparison.
  
```{r leukHepCgraph}
# Remove NAs in HepC_RNA variable for graphing purposes
dfadHepC2 <- dfad %>% drop_na(HepC_RNA)

ggplot(data=dfadHepC2, mapping=aes(x=HepC_RNA, y=Leukocytes, color=HepC_RNA)) +
  geom_boxplot() +
  xlab("Hepatitis C RNA") +
  ylab("Leukocytes (1000 cells/uL)") +
  stat_compare_means(method="aov")

ggsave("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/Leukocytes_by_HepCRNA.pdf")
```
  
### Results: Leukocyte concentrations by HepC RNA result
  
#### Leukocyte concentrations do not differ significantly based on hepatitis C RNA positivity status.
  
  
  
# Do lymphocyte and RBC levels correlate with one another?
  
Let's do a Pearson correlation to address this question.
  
```{r LymphRBC}
# Pearson correlation 
cor.test(x=dfad$Lymphocytes, y=dfad$RBCs, method="pearson")
```
  
Let's make a graph to visualize this comparison.
  
```{r LymphRBCgraph}

ggplot(data=dfad, mapping=aes(x=Lymphocytes, y=RBCs)) +
  geom_point() +
  xlab("Lymphocytes (%)") +
  ylab("Red Blood Cells (million cells/uL)") +
  geom_smooth(method="lm")
  stat_cor(method="pearson")

ggsave("C:/Users/mecho/Documents/Emory/Post-doc/Presentations/IntroToClinicalResearch_GuestLecture/Lymphocytes_by_RBCs.pdf")
```
  
### Results: Lymphocyte and RBC concentrations
  
#### Lymphocyte and RBC concentrations are not correlated with one another.
  
  
  
  





